Imperative Programming with Variational Effects

### 2018-08-13 paper/paper.tex
almost done

\title{Imperative Programming with Variational Effects}
\author{Alex Grasley}
\degree{Master of Science}
\doctype{Thesis}
\department{Electrical Engineering and Computer Science}
\depttype{School}
\depthead{Director}
\major{Computer Science}
\advisor{Eric Walkingshaw}
\submitdate{August 27, 2018}
\commencementyear{2018}
\abstract{
Variation is a commonly encountered element of modern software systems.
Recent research into variation has led to increasing interest in the
development of variational programming languages and corresponding
variational models of execution. Variational imperative languages pose a
particular challenge due to the complex interactions between side effects and
variation. The development of interpreters for variational imperative languages
has produced a number of techniques to address these interactions. This
thesis builds upon and formalizes these techniques by defining a formal
operational semantics for a simple imperative language that supports both
variation and common side effects.

@@

@@

This thesis has defined a formal operational semantics for a simple variational imperative language. It is our hope that
this could serve as the basis for a formal proof that VIMP satisfies the correctness property outlined in Section \ref{sec:props}.
Such a proof would give the highest level of confidence that the techniques outlined herein are sound and can safely be
applied to more complex languages.

@@

>>>>

\title{Imperative Programming with Variational Effects}
\author{Alex Grasley}
\degree{Master of Science}
\doctype{Thesis}
\department{Electrical Engineering and Computer Science}
\depttype{School}
\depthead{Director}
\major{Computer Science}
\advisor{Eric Walkingshaw}
\submitdate{August 27, 2018}
\commencementyear{2019}
\abstract{
Variation is a commonly encountered element of modern software systems.
Recent research into variation has led to increasing interest in the
development of variational programming languages and corresponding
variational models of execution. Variational imperative languages pose a
particular challenge due to the complex interactions between side effects and
variation. The development of interpreters for variational imperative languages
has produced a number of techniques to address these interactions. This
thesis builds upon and formalizes these techniques by defining a formal
operational semantics for a simple imperative language that supports both
variation and common side effects.

@@

The first to develop a notion of variational execution were Erwig and Walkingshaw in their 2011 GTTSE tutorial \cite{erwig2013variation}.
While it does not address the interaction of side effects with variation, it set the stage for future research by developing the choice calculus,
as well as the functor, applicative, and monad instances that we have made use of in this work.

@@

This thesis has provided a formal operational semantics for the variational imperative language VIMP.
The techniques formalized in VIMP were then demonstrated in the implementation of the Resource DSL.
Our intention is for this work to serve as the basis for future development of variational imperative languages.
By providing a formal semantics as well as a detailed description of the various techniques involved in the
implementation of variational imperative languages, we hope that future language developers will be able to
more fully focus on language-specific features, rather than being bogged down in the common details of the interaction
of variation with side effects.

It is our hope that the formal semantics that we have developed here
could serve as the basis for a formal proof that VIMP satisfies the correctness property outlined in Section \ref{sec:props}.
Such a proof would give the highest level of confidence that the techniques outlined herein are sound and can safely be
applied to more complex languages.

@@

For the Resource DSL, we are interested in the possibilities that built-in support for SAT solving has for optimization.
In our implementation of the test scenarios for the Resource DSL we noted that many computations that currently occur
at runtime in our language could potentially be offloaded to a SAT or SMT solver. We are also interested in continuing to
explore the domain of resource specifications for the potential to support fully declarative specifications. In the current iteration
of the Resource DSL we reluctantly included models as a way to encode fully imperative programs with support for sequencing
and control structures. If possible, we would like to move further in the direction of our declarative profiles. This however would require
ways of addressing resource configurations in which the order in which effects are performed and sequenced is important in a declarative
way, which ultimately may not be possible.
